---
title: "RNA-seq Preprocessing and Normalization"
author: "Zoe Dellaert"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    html_preview: false
---

# Preproccessing of bulk RNA-seq data

```{r setup}
# set up file paths so that Rmd outputs can be viewed using github markdown
knitr::opts_knit$set(base.dir = normalizePath("output_RNA/reports/"), base.url = "./")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,fig.width = 10, fig.height = 8,
                      fig.path = "DE_Analysis_files/figure-gfm/")

#load packages
library(tidyverse)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(genefilter)
library(ggnewscale)
library(BiocParallel)

# set number of cores to use for parallel DESeq2 processing
register(MulticoreParam(workers = 18))

sessionInfo() #provides list of loaded packages and version of R
```

```{r}
save_ggplot <- function(plot, filename, width = 10, height = 7, units = "in", dpi = 300,bg = "white") {
  png_path <- file.path(outdir_plots, paste0(filename, ".png"))
  pdf_dir <- file.path(outdir_plots, "pdf_figs")
  pdf_path <- file.path(pdf_dir, paste0(filename, ".pdf"))
  
  # Ensure the pdf_figs directory exists
  if (!dir.exists(pdf_dir)) dir.create(pdf_dir, recursive = TRUE)
  
  # Save plots
  ggsave(filename = png_path, plot = plot, width = width, height = height, units = units, dpi = dpi,bg = bg)
  ggsave(filename = pdf_path, plot = plot, width = width, height = height, units = units, dpi = dpi,bg = bg)
}
```


## 0. Setup species-specific parameters

```{r get-species}
# set up necessary output directories if they don't exist
outdir <- file.path("output_RNA/analysis")
outdir_plots <- file.path(outdir,"plots")
if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
if (!dir.exists(outdir_plots)) dir.create(outdir_plots, recursive = TRUE)

reportdir <- file.path("output_RNA/reports/DE_Analysis_files/figure-gfm/")
if (!dir.exists(reportdir)) dir.create(reportdir, recursive = TRUE)
```

## 1. Read in raw count data

```{r load-counts}
# load in data
counts_raw <- read.csv(file.path("output_RNA/count_matrices/POR_Pcomp_gene_count_matrix.csv"), row.names = 1)

# make list of samples 
samples <- colnames(counts_raw)
cat("Raw counts:", nrow(counts_raw), "genes x", ncol(counts_raw), "samples")

# read in SwissProt annotation
SwissProt <- read.delim(file.path("../HI_genome_annotations/annotation/Porites_compressa_HIv1_Swissprot_GO.tsv"))
cat("Annotations:", nrow(SwissProt), "Swissprot-annotated genes")
```

## 2. Extract metadata from sample names

```{r metadata}
# remove _S* from sample names
colnames(counts_raw) <- gsub("\\_S\\d\\d","",colnames(counts_raw))
samples <- colnames(counts_raw)

# create metadata dataframe from sample names
meta <- data.frame(
  sample = samples, 
  slide = c("POR_3_Slide1","POR_4_Slide3","POR_R72_C2_1"),
  tissue = c("bulk","bulk","Oral_Gastro")
)

# add rownames
rownames(meta) <- meta$sample

# make time and treatment factors
meta$tissue <- factor(meta$tissue)

# save metadata
write.csv(meta, paste0("output_RNA/RNA_seq_metadata.csv"))

# reorder count matrix to be in order of metadata table (should be already but just in case)
counts_raw <- counts_raw[, meta$sample]
```

## 3. Remove outliers, if identified

Nope!

## 4. pOverA filtering to reduce dataset

### Note to self: maybe replace this with treatment-specific filtering. To get genes expressed only at one timepoint in one treatment

```{r filtering}
# Keep genes expressed at 10+ counts in at least 33% of samples (aka one sample)

ffun<-filterfun(pOverA(0.33,10))
counts_filt_poa <- genefilter((counts_raw), ffun) #apply filter

filtered_counts <- counts_raw[counts_filt_poa,] #keep only rows that passed filter

paste0("Number of genes after filtering: ", sum(counts_filt_poa))
paste0("% of genes kept: ", round(100*(sum(counts_filt_poa)/nrow(counts_raw)),digits=2),"%")

write.csv(filtered_counts, file = file.path(outdir, "filtered_counts.csv"))
```

## 5. Create DESeq object and run DESeq2

```{r deseq}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ tissue)

dds <- DESeq(dds, parallel = TRUE)

# Estimate size factors to determine if we can use VST
SF.dds <- estimateSizeFactors(dds) 
print(sort(sizeFactors(SF.dds))) #View size factors

# if all are less than 4 we can use the VST transformation
all(sizeFactors(SF.dds)) < 4
```

## 6. VST-Transforming count data for visualization 

```{r vst-transform}
vst <- vst(dds, blind=FALSE)

#save the vst transformation
vst_mat <- assay(vst)
write.csv(vst_mat, file = file.path(outdir, "vst_expression_matrix.csv"))
```

## 7. Visualize sample-sample relationships

### PCA

```{r pca}
pcaData <- plotPCA(vst, intgroup=c("tissue"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

PCA_simple <- ggplot(data = pcaData, aes(x=PC1, y=PC2, color=tissue, shape=sample)) +
  geom_point(size=4) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  labs(color = "Tissue", shape = "Sample") +
  coord_fixed() + theme_bw() + ggtitle("PCA of VST-transformed counts")

print(PCA_simple)
save_ggplot(PCA_simple, "PCA_simple", width = 8, height = 6)
```

### Hierarchical Clustering

```{r cluster}
sampleTree <- hclust(dist(t(vst_mat)), method = "average")

par(mar = c(8, 4, 2, 2))
plot(sampleTree, 
     xlab = "", sub = "", cex = 0.7)
abline(h = 100, col = "red", lty = 2)
```

###  Heatmap of variable genes

```{r heatmap}
topVarGenes <- head(order(rowVars(vst_mat), decreasing=TRUE), 500)

png("output_RNA/analysis/plots/top500vargenes_heatmap.png")
pheatmap(vst_mat[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta %>% select(tissue))
dev.off()

```

### Text summary

```{r final, echo=FALSE}
cat("Preprocessing Summary:")

cat("Input\n")
cat(strrep("-", 40), "\n")
cat("  Initial genes:", nrow(counts_raw), "\n\n")

cat("Filtering\n")
cat(strrep("-", 40), "\n")
cat("  Low-expression genes removed:", sum(!counts_filt_poa), "\n")
cat("  pOverA filter: >=", 10, "counts in >=", 
    0.33 * 100, "% of samples\n\n")

cat("Output\n")
cat(strrep("-", 40), "\n")
cat("  Final genes:", nrow(vst_mat), "\n")
cat("  Final samples:", ncol(vst_mat), "\n")
cat("  Output directory:", outdir, "\n\n")

cat("QC Notes\n")
cat(strrep("-", 40), "\n")
cat("  Size factors range:", round(min(sizeFactors(SF.dds)), 2), "-", round(max(sizeFactors(SF.dds)), 2), "\n")
cat("  VST appropriate:", ifelse(all(sizeFactors(SF.dds) < 4), "Yes", "No - consider rlog"), "\n")
cat("  PC1 variance:", percentVar[1], "%\n")
cat("  PC2 variance:", percentVar[2], "%\n")
```

```{r}
sessionInfo()
```

